\chapter*{History}

The theory of general relativity was being developed by Albert Einstein between
1907 and 1915. It builds upon and combines the earlier theories of Newtonian
gravity and special relativity. Although the basic theory hasn't changed since
then, there have been many contributions afterwards. The following timeline
summarizes the history as far it is relevant to the notes at hand.

\begin{description}
\item[1609]
  Kepler pushlishes his first two laws of planetary motion.
\item[1619]
  Kepler pushlishes his third law of planetary motion.
\item[1638]
  Galilei's equations for a falling body.
\item[1687]
  Newton publishes his law of universal gravitation.
\item[1798]
  Cavendish measures Newton's gravitational constant.
\item[1862]
  Maxwell's equations of electromagnetism.
\item[1887]
  The Michelson-Morley experiment fails to detect a stationary
  luminiferous aether.
\item[1889]
  FitzGerald proposes Lorentz contraction.
\item[1905]
  Formulation of special relativity by Einstein.
\item[1915]
  Derivation of Einstein's field equations from an action principle by Hilbert.
\item[1915]
  Einstein's theory of general relativity.
\item[1916]
  Schwarzschild found the first exact solution of Einstein's field equations.
\item[1916]
  Einstein shows that the perihelion precession of Mercury can be fully
  explained by general relativity.
\item[1919]
  Eddington's expedition confirms that the deflection of light by the Sun is as
  predicted by general relativity.
\item[1922]
  Friedmann found a cosmological solution to Einstein's equations, in which the
  universe may expand or contract.
\item[1922]
  Introduction of the cosmological constant by Einstein into his field equations.
\item[1922]
  Proposal of the Einstein--Cartan theory by Élie Cartan.
\item[1929]
  Hubble finds evidence that the universe is expanding.
\item[1959]
  Direct measurement of the gravitational redshift of light in the Pound-Rebka
  experiment.
\item[1964]
  Discovery of the cosmis microwave background by Penzias and Wilson.
\item[1964]
  Discovery of the X-ray source Cygnus X-1, now widely accepted to be a black
  hole.
\end{description}
